---
title: "Faunalytics R Tables and Figures Style Guide"
output: html_document
date: "2024-10-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(faunalytics)
library(gt)
```

# Requirements

## Install CRAN packages
Make sure you have installed both the `tidyverse` (which contains ggplot2,
dplyr, and many other common packages) and the `gt` package (tables). You will 
also need the `devtools` package.

```{r install-cran, eval = FALSE}
install.packages("tidyverse")
install.packages("gt")
install.packages("devtools")
```

## Install the `faunalytics` package
Install the most recent version of the `faunalytics` package:
```{r install-fauna, eval = FALSE}
devtools::install_github("faunalytics/faunalytics")
```

## Install and load the Gotham font family
If you have not already, make sure to download and install all of [the Gotham
fonts](https://drive.google.com/file/d/1ZgXfzdITIsMVlqZKRcLKiERzznSxfhDq/view?usp=sharing).

Once installed, they can be loaded in R using the extrafont package.
```{r load-gotham, eval = FALSE}
install.packages("extrafont") # install extrafont package
extrafont::font_import(pattern = "Gotham") # load Gotham fonts
extrafont::loadfonts(device = "win") # only needed if you're on a Windows computer
```

If you run into issues getting the fonts to work, there are a lot of
troubleshooting resources online. You can also contact Zach for support.

The only functions (as of October 28, 2024) that directly call on the Gotham
font family are `faunalytics::theme_faunalytics()` and 
`faunalytics::table_format()`. Both of these also take an argument `gotham` that
is `TRUE` by default, but can be set to `FALSE`, which will make it use
Helvetica instead of Gotham. For example, `theme_faunalytics(gotham = FALSE)`
or `table_format(gotham = FALSE)`.

## Webshot
We use the `gt` package for tables. In order to save `gt` objects to PNGs like
we want to, we'll need to install the `webshot` package and use it to install
PhantomJS. Like all R installations, you'll only need to do this once.
```{r web-phantom, eval = FALSE}
install.packages("webshot")
webshot::install_phantomjs()
```

# Overview

***An important note:*** You should default to the styles and standards outlined in this
document. However, there may be cases when it makes more sense to deviate from 
these guidelines for the sake of clarity and/or readability.

As communicators of scientific information, it is important that our work is 
clear and consistent. This is true for both our writing and our visuals. To that
end, there are certain styles and standards we should use to make sure that we
have a consistent aesthetic across reports.

The `faunalytics` package comes with several functions designed to make it 
easy to follow these standards. These include `theme_faunalytics` for `ggplot`
objects and `table_format` for `gt` tables, both of which are discussed below.
Other helpful functions include `fauna_colors` and `fauna_blues`, also discussed
below. 

## Colors

The standard Faunalytics color palette includes green, amber, red, and blue,
which can be broken down into dark blue, light blue, and regular blue. Rather 
than the standard shades of these colors (e.g., #FF0000 for red), we use
specific brand shades (e.g., #E64B3D for red). The specific shades we use can
be accessed using the `fauna_colors` function.

```{r colors}
# return hex code for red
fauna_colors("red")
# return all hex codes in standard palette
fauna_colors()
# return all hex codes in standard palette with names for clarity
fauna_colors(nameless = FALSE)
```
```{r color-plot, echo=FALSE}
scales::show_col(fauna_colors())
```

Unfortunately, some of these colors don't work well together (e.g., green and
red aren't colorblind-friendly). Green is also often associated with "good" and
red with "bad." To avoid issues and limitations such as these, graphics should
generally default to shades of blue. 

When using only one color, use dark blue (#254C59) . When using two colors, use 
dark blue and light blue (#5FB7E5). When using three colors, use dark blue, 
regular blue (#0092B2), and light blue. If appropriate, use additional shades
of blue when using more than three colors. You can access all of these shades
of blue by providing the number of shades you need to the `fauna_blues` 
function. When the number you provide is greater than three, the function
interpolates shades between dark blue and light blue.

```{r blues}
fauna_blues(1) # dark blue only
fauna_blues(2) # dark blue and light blue
fauna_blues(3) # dark blue, regular blue, and light blue
fauna_blues(6) # dark blue, four interpolated shades, and light blue 
```

# ggplot

We typically use the `ggplot2` package for graphs. At the end of your ggplot
chunk, you can add `+ theme_faunalytics()` to standardize many aspects of your
plot's theme. For example, notice the difference between these two plots below.

We'll summarize the mtcars data for the examples below.

```{r mtcars-factor, echo = FALSE}
mtcars2 <- mtcars |> 
  group_by(cyl) |> 
  summarize(mpg = mean(mpg)) |> 
  ungroup() |> 
  mutate(cyl = factor(cyl, levels = c(4, 6, 8)))
```

```{r theme_faunalytics-example}
ggplot(mtcars2, aes(x = cyl, y = mpg)) +
  geom_col() # without the Faunalytics theme

ggplot(mtcars2, aes(x = cyl, y = mpg)) +
  geom_col() +
  theme_faunalytics() # with the Faunalytics theme
```

***Note:*** In `ggplot`, if you're referring to a variable in your dataset,
it should be referenced within the `aes()` command, like `aes(x = cyl, y = mpg)`.
However, if you're referencing something that's *not* in your data, it should
be referenced outside of the `aes()` command, like `fill = fauna_colors("darkblue")`.

Where white would be used, we just stick with the standard white (#FFFFFF). You
can also access this by using `fauna_colors("white")`. However, rather than
black, we use a slightly softer dark gray (#333333), accessed using
`fauna_colors("darkgray")`.

You may notice that the graph is not entirely in the Faunalytics colors scheme. 
Unfortunately, this can't be automatically applied using `theme_faunalytics` and
requires specification. For example:
```{r make-plot-blue1}
ggplot(mtcars2, aes(x = cyl, y = mpg)) +
  geom_col(fill = fauna_blues(1)) + # make all columns the same color
  theme_faunalytics()
```

or
```{r make-plot-blue2}
ggplot(mtcars2, aes(x = cyl, y = mpg)) +
  geom_col(aes(fill = cyl)) + # change color based on cyl value
  scale_fill_manual(values = fauna_blues(3)) +
  theme_faunalytics()
```

*Note:* In ggplot terms, `fill` is the term for bars and other things that are
*filled in* with a color. `color` refers to lines, points, and borders. 

## Maximizing the data-ink ratio

In general, we want to try to adhere to statistician and data viz legend [Edward
Tufte's principles for graphing](https://thedoublethink.com/tuftes-principles-for-visualizing-quantitative-information/),
which emphasize showing the data and reducing the amount of unnecessary
components on a graphic (aka maximizing the data-ink ratio). To this end, we
often want to remove things like axis titles, axis text, axis ticks, etc. if
they aren't providing critical information that isn't available somewhere else. 
Take the following graph as an example. The graph title appears before the code
because we type this titles into reports rather than embedding them in the
graph images. 

#### Average miles per gallon by number of engine cylinders
```{r max-plot}
ggplot(mtcars2, aes(x = cyl, y = mpg)) +
  geom_col(aes(fill = cyl)) +
  scale_fill_manual(values = fauna_blues(3)) +
  theme_faunalytics()
```

The y axis title and x axis title are made redundant by the graph title. The 
x axis text could also be eliminated if we more clearly label the legend. Since
we won't need the x axis text, we can also eliminated the associated ticks. We 
can get rid of this redundant information using a `theme()` command, plus an
additional tweak to rename the legend. 

#### Average miles per gallon by number of engine cylinders
```{r min-plot}
ggplot(mtcars2, aes(x = cyl, y = mpg)) +
  geom_col(aes(fill = cyl)) +
  scale_fill_manual("Engine Cylinders", # rename legend
                    values = fauna_blues(3)) +
  theme_faunalytics() +
  theme(axis.title = element_blank(), # hide axis titles
        axis.text.x = element_blank(), # hide x axis text (i.e., 4, 6, 8)
        axis.ticks.x = element_blank()) # hide x axis ticks
```

***Note: *** The `theme` command must go *after* the `theme_faunalytics` command.

You can get rid of elements by setting them equal to element_blank() as done
in the above example. For additional ways to change the defaults, see [this 
ggplot2 guide](https://ggplot2.tidyverse.org/reference/theme.html). 

While we're at it, we can make a few more tweaks to make the graphic better.

#### Average miles per gallon by number of engine cylinders
```{r min-plot2}
ggplot(mtcars2, aes(x = cyl, y = mpg)) +
  geom_col(aes(fill = cyl)) +
  scale_fill_manual("Engine Cylinders", 
                    values = fauna_blues(3)) +
  scale_y_continuous(limits = c(0, 30), # set y axis limits
                    breaks = seq(0, 30, 10)) + # set y tick locations (0 to 30 every 10 units))
  guides(fill = guide_legend(title.position = "top")) + # put legend title above legend
  theme_faunalytics() +
  theme(axis.title = element_blank(), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) 
```
  
## Data labels
We can also add data labels. This may seem like it contradicts the data-ink
ratio maximization guideline, but it doesn't really because we're using these 
data labels to provide additional information. Rather than guessing what the
heights of the bars are, we're providing additional information.

```{r add-text-plot2}
ggplot(mtcars2, aes(x = cyl, y = mpg)) +
  geom_col(aes(fill = cyl)) +
  geom_text(aes(label = sprintf("%.1f", mpg), # label with mpg, rounded to 1 decimal place
                y = mpg + 1), # place label one unit above top of bar
            color = fauna_colors("darkgray"), # set text to dark gray
            family = "Gotham Bold") + # change font to Gotham Bold
  scale_fill_manual("Engine Cylinders", 
                    values = fauna_blues(3)) +
  scale_y_continuous(limits = c(0, 30),
                    breaks = seq(0, 30, 10)) + 
  guides(fill = guide_legend(title.position = "top")) +
  theme_faunalytics() +
  theme(axis.title = element_blank(), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) 
```

Above, we change the font to Gotham Bold. This is our standard **bold** font.
For non-bold fonts, we use Gotham Book. We also changed the default text color
(black) to dark gray and moved the label up a bit so it wouldn't overlap the bar.
These are all adjustments to how our `label` shows up. We could have simply set 
this to `mpg`, but it would have given a lot of decimal places. Since we only 
want one decimal place, we wrap it in the `sprintf` function. "%.1f" gives us 
one decimal place (.1 for one decimal place, .2 for two decimal places, etc.). 
We could have use `round(mpg, 1)` instead, but this would not have displayed a 0
in the decimal place if any existed. 

## Other things

### Error bars
We set the color of error bars to gray (#808080) so they show up clearly. You
can do this using `fauna_colors("gray")`. We also set the default width of
error bars to .15. 

## Grouped bars, labels, and error bars
If you want to make a grouped bar plot, you have to tell `ggplot` that by setting
the fill to a variable. This will create a *stacked* bar plot. To make it a 
grouped bar plot, you have to tell `ggplot` to unstack it. To do this, we add 
the argument `position = position_dodge()` to the `geom_*` line. By default,
the width (distance of the grouped bars) is .9. If you have labels on the bars
or error bars, you'll also need to set those to be position dodged and you'll
need to manually set the width to .9: `position = position_dodge(width = .9)`. 
You can see an example of all of this below using a different summarized version
of the `mtcars` dataset.

```{r mtcars3-maker, echo = FALSE, message = FALSE}
mtcars3 <- mtcars |> 
  group_by(cyl, am) |> 
  summarize(mean_mpg = mean(mpg),
            n = n(),
            se = sd(mpg) / sqrt(n),
            lower_ci = mean_mpg - qt(1 - (0.05 / 2), n - 1) * se,
            upper_ci = mean_mpg + qt(1 - (0.05 / 2), n - 1) * se
            ) |> 
  ungroup() |> 
  rename(mpg = mean_mpg) |> 
  mutate(cyl = factor(cyl, levels = c(4, 6, 8)),
         am = case_when(
           am == 0 ~ "Automatic",
           am == 1 ~ "Manual"
         ),
         am = factor(am, levels = c("Automatic", "Manual")))
```
#### Average miles per gallon by number of engine cylinders and transmission type
```{r error-bars-and-dodging}
ggplot(mtcars3, aes(x = cyl, y = mpg, 
                    fill = am)) + # move fill to ggplot line so it applies to everything below
  geom_col(position = position_dodge()) + # set position_dodge for bars (columns)
  geom_errorbar(aes(ymin = lower_ci, # set error bar minimum to pre-calculated value
                    ymax = upper_ci), # # set error bar maximum to pre-calculated value
                color = fauna_colors("gray"), # set error bar color to gray
                width = .15, # set error bar width
                position = position_dodge(width = .9)) + # specify dodged position and width
  scale_fill_manual("Tranmission Type", 
                    values = fauna_blues(2)) +
  scale_y_continuous(limits = c(0, 40),
                    breaks = seq(0, 40, 10)) +
  labs(x = "Number of Engine Cylinders") +
  guides(fill = guide_legend(title.position = "top")) +
  theme_faunalytics() +
  theme(axis.title.y = element_blank()) # hide only y axis title
```

### Axis scaling
Axes should also be appropriately scaled. For example, if we're showing
percentages and a 0%-100% axis, the axis shouldn't show up as 0.0 to 1.0, it 
should show up as 0% to 100%. (Data labels should match this as well.) 

The `scales` package allows us to make these changes fairly easily. Take the
following graph as an example:

#### Percentage of cars in `mtcars` data by number of cylinders
```{r bad-scales}
mtcars |>
 tabyl(cyl) |> 
  ggplot(aes(x = as.factor(cyl), y = percent)) +
  geom_col(fill = fauna_colors("darkblue")) +
  labs(x = "", y = "") + # Set axis titles to blank
  theme_faunalytics()
```

We're reporting percentages, so the y axis should reflect that more clearly. We
can set the `labels` argument of the `scale_y_continous` (because we want to 
adjust the y axis of a continuous value, as opposed to `scale_y_discrete`) equal
to `scales::percent` to tell ggplot to treat the y axis values as percentages.  
(The double colon is how you access a single function from an unloaded package:
`package::function`. If we had run `library(scales)` earlier, we could have just
written `percent`.)

#### Percentage of cars in `mtcars` data by number of cylinders
```{r better-scales}
mtcars |>
 tabyl(cyl) |> 
  ggplot(aes(x = as.factor(cyl), y = percent)) +
  geom_col(fill = fauna_colors("darkblue")) +
  scale_y_continuous(limits = c(0, .5), # set minimum and maximum of axis
                     labels = scales::percent) + # show axis labels as percentages
  labs(x = "", y = "") + 
  theme_faunalytics()
```

Similarly, we should format scales in the thousands and larger with commas.
Again, we can use the `scales` package with the `scale_y_continuous` function.

#### Car weight by miles per gallon
```{r comma-scales}
mtcars |>
 mutate(wt = wt * 1000) |> # convert wt from showing thousands of lbs to showing lbs
  ggplot(aes(x = mpg, y = wt)) +
  geom_point(color = fauna_colors("darkblue")) + # geom_point uses color, not fill
  scale_y_continuous(limits = c(0, 6000), # set axis min and max
                     labels = scales::comma) + # use commas in axis text
  labs(x = "Miles per gallon", y = "Weight (lbs)") + 
  theme_faunalytics()
```

The `scales` package comes with lots of ways to format axes, so it's worth
exploring it if you have/want to tweak axis labels.


## Saving
You can save images with the `ggsave` function. It safest to save your
graph as an object using the `<-` operator and then call that plot by name
(versus the alternative of relying on `ggsave`'s default "last plot" behavior). 

`ggsave` requires you to specify a filename you'd like to save your plot as. You
can include a full file path here and save the image as a png. Width and height
are up to you, but changing these can sometimes affect how the plot renders,
so be sure to check your saved image to make sure it looks correct.

An example of the ggsave syntax for a non-existent plot called `p_label_belief`:
```{r, eval = FALSE}
ggsave(filename = "output/figures/welfare-label-beliefs.png", 
       plot = p_label_belief)
```

# Tables

Tables are annoying unfortunately. We incorporate a package called `gt` to make
consistent, save-able tables. You can apply the default formatting using the
function `table_format` from the `faunalytics` package. This function takes
a dataframe and makes it a `gt` table. It will automatically name the columns
whatever they're name in your dataframe, so you may wish to rename them before
using the `table_format` function.

Before showing this, let's make some data to work with.
```{r tbl-dat}
mtcars_auto_tbl <- mtcars |> 
  filter(am == 0) |> # subset data to keep only automatic transmission cars
  rownames_to_column("car") |> # turn rownames into a column called "car'
  mutate(make = trimws(str_extract(car, "^.*?\\s|^Valiant$"))) |> # extract car make
  tabyl(make) # create summary table
```

```{r gt-example}
mtcars_auto_tbl |> 
  table_format()
```

By default, any table with four or more rows gets alternate row shading to make
the tables easier to read. This and many other features are changeable. See
`?table_format` for more on that. 

Unfortunately, many features have to be changed using `gt`'s clunky syntax. For 
anything not covered here, have a look at [this guide to gt](https://gt.rstudio.com/articles/gt.html).

## Alignment and column width

We'll need to change the names of the columns, but let's save that for last so
that we don't have to deal with capitalization and so on. 

Unlike with `ggplot`, it's tricky to reformat columns in `gt`, so we'll reformat
our percentages and numbers before applying the `table_format` function. 

We'll want to right-align all numeric columns. In this case, that's our `n` and
`percent` columns. You can do that using `gt`'s `cols_align` function like this:
```{r gt-nums}
mtcars_auto_tbl |>
  mutate(percent = paste0(round(100 * percent, 0), "%"), # reformat percentage
         n = format(n, big.mark = ",")) |> # add comma separators
  table_format()
```

The line adding comma separators isn't necessary because our numbers are all
below 1,000, but I've included it because it could be helpful in other cases.

We might get the proper alignment for our columns, but just in case we didn't
or we want to be extra safe, we can specify column alignment like this:
```{r gt-align}
mtcars_auto_tbl |>
  mutate(percent = paste0(round(100 * percent, 0), "%"),
         n = format(n, big.mark = ",")) |>
  table_format() |> 
  cols_align(align = "right", # specify alignment direction
             columns = c(n, percent)) # name columns in vector (using c())
```

We can specify column width as well. You can do this with either pixels (`px`)
or with percentages (`pct`). `px` is probably the better option in most cases.
As a note, the `cols_width` functionworks with tidyverse shortcut functions like
`contains` or `starts_with`. 

```{r gt-width}
mtcars_auto_tbl |>
  mutate(percent = paste0(round(100 * percent, 0), "%"),
         n = format(n, big.mark = ",")) |>
  table_format() |> 
  cols_align(align = "right",
             columns = c(n, percent)) |> 
  cols_width(make ~ px(100), # specify width in pixels
             n ~ px(110), # bigger than needed here so we can rename it
             percent ~ px(110))
```

We can also format specific rows or columns. For example, let's say we wanted
a "Total" row at the bottom of this table. We'd want to make a border line
and to bold the text. This is where things get extra clunky. You'll need to use
`gt`'s tab_style function. See the guide linked above or `?gt::tab_style` for 
more info on how to use this.

```{r gt-total}
mtcars_auto_tbl |>
  adorn_totals() |> # add totals row to tabyl object
  mutate(percent = paste0(round(100 * percent, 0), "%"),
         n = format(n, big.mark = ",")) |>
  table_format() |> 
  cols_align(align = "right",
             columns = c(n, percent)) |> 
  cols_width(make ~ px(100),
             n ~ px(110),
             percent ~ px(110)) |> 
    tab_style( # format specific cells
    style = list(
      cell_text(weight = "bold"), # make text bold
      cell_borders(sides = c("top"), # add border to top of specific cells
                   color = fauna_colors("darkgrey"), # make border dark gray
                   weight = px(2)) # make border 2px thick
    ),
    locations = cells_body(
      rows = percent == "100%" # apply the above style to rows where the
      # value of the percent column is equal to "100%"
    ))
```

Now we can rename the columns to finalize the table.

```{r gt-rename}
mtcars_auto_tbl |>
  adorn_totals() |>
  mutate(percent = paste0(round(100 * percent, 0), "%"),
         n = format(n, big.mark = ",")) |>
  table_format() |> 
  cols_align(align = "right",
             columns = c(n, percent)) |> 
  cols_width(make ~ px(100),
             n ~ px(110), 
             percent ~ px(110)) |> 
    tab_style(
    style = list(
      cell_text(weight = "bold"), 
      cell_borders(sides = c("top"), 
                   color = fauna_colors("darkgrey"), 
                   weight = px(2))
    ),
    locations = cells_body(
      rows = percent == "100%"
    )) |> 
    cols_label(make = "Car Make",
               n = "Frequency",
               percent = "Percentage")

```

In case we wanted to make a linebreak between the words "Car" and "Make",
we could replace the space with "<br/>" and wrap the whole thing in `md()`, 
which is short for "markdown", the style of formatting. This lets `gt` know to
evaluate "<br/>" as a line break rather than just treat it like normal text.

```{r gt-break}
mtcars_auto_tbl |>
  adorn_totals() |>
  mutate(percent = paste0(round(100 * percent, 0), "%"),
         n = format(n, big.mark = ",")) |>
  table_format() |> 
  cols_align(align = "right",
             columns = c(n, percent)) |> 
  cols_width(make ~ px(100),
             n ~ px(110), 
             percent ~ px(110)) |> 
    tab_style(
    style = list(
      cell_text(weight = "bold"), 
      cell_borders(sides = c("top"), 
                   color = fauna_colors("darkgrey"), 
                   weight = px(2))
    ),
    locations = cells_body(
      rows = percent == "100%"
    )) |> 
    cols_label(make = md("Car<br/>Make"), # add line break and markdown specification
               n = "Frequency",
               percent = "Percentage")
```

To save this, we'd store it as an object using `<-` and run the appropriate
save function, just like with a `ggplot` graph. In this case, we'd use `gtsave`.
On a non-existent object called `gt_tbl`, we'd save that like this:
```{r gt-save, eval = FALSE}
gtsave(gt_tbl, "output/tables/table-1-results.png") # save table as png
```
As noted above, this requires webshot2 to be installed. 

# TO ADD:
- Gotham installation
- need to do something about reproducability without Gotham
    - perhaps: build into the two Gotham-using pckgs a check. if no Gotham,
      then WORK ANYWAY but throw a warning. also give option to avoid warning
      by specifying gotham = FALSE
- webshot installation
- faunalytics package updates
- why do i have to load gt?
- continuous color ramp